---
title: "<span style='font-size: 27px'>Analysis of 2022 Indiana Terminated Pregnancy Reports </span>"
author: "Ted Schurter"
date: "2023-07-14"
output:
  html_document:
    
theme: lumen
---

<style type="text/css">
.main-container {
  max-width: 1600px;
  margin-left: autopx;
  margin-right: 200px;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(echo = F)
knitr::opts_chunk$set(warning = F)
knitr::opts_chunk$set(message = F)
knitr::opts_chunk$set(cache = F)
```

```{r}
library(tidyverse)
library(readxl)
library(ggtext)
library(ggrepel)
library(DT)
library(patchwork)
library(tigris)
```


```{r, set ggplot theme}
# set custom theme ####

# panel color 
panel_c <- "#fdfdf2"

# theme
t_theme <- function(base_size = 10){
  
  theme_classic() %+replace%
    
    theme( 
      #grid elements
      panel.grid.major.y = element_line(color = "gray85", size = .25),
      #panel.grid.major.y = element_blank(),
      panel.grid.major.x = element_blank(), 
      axis.line = element_blank(),
      axis.ticks = element_blank(), 
      
      #axis.line = element_line(color = "gray90"),
      #axis.line.y = element_blank(),
      
      axis.text = element_text(color = "gray55",
                               size  = rel(.65)),
      axis.title.x = element_blank(),
      # element_text(color = "gray55",
      # size = rel(.65), hjust = 0),
      axis.title.y = element_blank(),
      # element_text(color = "gray55",
      #             size = rel(.6)),
      
      plot.title = element_textbox_simple(size = rel(1.5), color = "gray25", 
                                          hjust = 0,lineheight = 1, 
                                          margin = margin(0.2, 0, .2, 0, unit = "cm"),
                                          family = "serif",face = "plain"),
      plot.subtitle = element_textbox_simple(size = rel(1.1), color = "gray40", hjust = 0,
                                             lineheight = 1,margin = margin(0.1, 0, .4, 0, unit = "cm"),),
      plot.caption = element_textbox_simple(family = "sans", size = rel(.7),
                                            color = "gray40", halign = 1,
                                            lineheight = 1.2, margin = margin(0, 0, .1, 0, unit = "cm"),),
      plot.title.position = "plot",
      plot.caption.position = "panel", 
      plot.margin = margin(.5, .5, .5,.5, unit = "cm"),
      plot.background = element_rect(fill  = panel_c,
                                     color = panel_c),
      panel.background = element_rect(fill = panel_c,
                                      color = panel_c)
    )
}
```

```{r, import year by month data}
# monthly total data
mnth <- read_csv("exported_data/monthly_abortions_14_22.csv")

# order months as factor
mnth$month <- factor(mnth$month, levels = c(
  "January", "February", "March", "April", "May", "June", "July","August",
  "September", "October", "November", "December"))

# provider data
ploc <- read_csv("exported_data/provider_residency.csv")

# residency data
nres <- read_csv("exported_data/nonresident_ab_19_22.csv")

```

#### Total abortions in Indiana rose `r round(100*(sum(mnth$tot[mnth$year == 2022])-sum(mnth$tot[mnth$year == 2021]))/sum(mnth$tot[mnth$year == 2021]))`% to `r formatC(sum(mnth$tot[mnth$year == 2022]), big.mark = ",")` in 2022, boosted by a `r round(100*((sum(mnth$nonres[mnth$year == 2022])-sum(mnth$nonres[mnth$year == 2021]))/sum(mnth$nonres[mnth$year == 2021])))`% increase in nonresident abortions from women in neighboring states fleeing state abortion bans implemented following the elimination in June 2022 of federal abortion protection by the [Supreme Court](https://www.supremecourt.gov/opinions/21pdf/19-1392_6j37.pdf). Abortions to Indiana residents declined `r round(abs(100*((sum(mnth$res[mnth$year == 2022])-sum(mnth$res[mnth$year == 2021]))/sum(mnth$res[mnth$year == 2021]))))`% even as the impact of the state's own ban was [blunted](https://apnews.com/article/abortion-health-indiana-legislature-9dda020bf811f19e6254915ca1324edc) by a lawsuit that preserved abortion access while litigation continued.

#### While the Indiana Supreme Court recently [upheld](https://apnews.com/article/indiana-abortion-ban-dc149c0716e3cf5be604bb5f427db7b8) that ban, the impact of the shifting legal landscape in and around Indiana is reflected in 2022's volatile monthly abortion totals. Although they climbed above last year’s levels in February and remained there through September, they spiked much higher in July, the first month after the Dobbs decision ended federal abortion protection, reaching `r formatC(mnth$tot[mnth$year == 2022 & mnth$month == "July"], big.mark = ",")`, `r round(100*(mnth$tot[mnth$year == 2022 & mnth$month == "July"]- mnth$tot[mnth$year == 2021 & mnth$month == "July"])/ mnth$tot[mnth$year == 2021 & mnth$month == "July"])`% higher than July 2021.

## {.tabset}

### Chart

```{r, year by month plot, fig.width=11, fig.height=6.5, fig.alt= "Line graph of total monthly abortions from 2014 to 2022 showing record high numbers in July 2022 and record low numbers in October 2022. Chart also shows significant dates in 2022 related to abortions changing legal status at the federal and Indiana level."}
# set colors 

# 2021 color 
c_2021 <- "#1b7837"

# 2022 color 
c_2022 <- "#762a83"

# assign colors for labeling
c_mnth <- mnth %>% filter(month == "December") %>% mutate(
  color = 
    case_when(year == 2022 ~ c_2022,
              year == 2021 ~ c_2021, 
              .default = "black")
)

# plot

ggplot()+ 
  # lines for all months
  geom_line(data = mnth %>% group_by(year) %>% arrange(desc(tot)),
            aes(month, tot, group = year), color = "gray85", size = .25)+
  
  # Dobbs leak text and marker 
  # vert line
  geom_segment(aes(x=5+(2/31), xend = 5+(2/31),
                   y = 200, yend = 1200),
               color = "gray80", linetype = "longdash", linewidth = .15)+
  # label
  geom_textbox(
    aes(x= 5, y = 200,
        label = "**May 2, 2022:**<br><br>Draft of US Supreme Court Dobbs v. Jackson 
        decision is leaked, raising speculation about overturn of Roe v. Wade."),
    label.colour = NA, fill = panel_c,
    size = rel(2.65), color = "gray50", 
    hjust = 1, vjust = 0, box.color = NA, width = unit(.9, "inch"))+
  
  # Dobbs text and marker line
  # vert line
  geom_segment(aes(x=6+(24/30), xend = 6+(24/30),
                   y = 200, yend = 1200),
               color = "gray80", linetype = "longdash", linewidth = .15)+
  # label 
  geom_textbox(
    aes(x= 6+(24/30)-.01, y = 200,
        label = "**June 24, 2022:**<br><br>Dobbs decision overturning 
        Roe v. Wade and removing constitutional right to abortion is released."),
    label.colour = NA, fill = panel_c,
    size = rel(2.65), color = "gray50", 
    hjust = 1, vjust = 0, box.color = NA, width = unit(1.15, "inch"))+
  
  # Aug. 5 text and marker line
  # vert line
  geom_segment(aes(x=8+(5/31), xend = 8+(5/31),
                   y = 200, yend = 1200),
               color = "gray80", linetype = "longdash", linewidth = .15)+
  # label
  geom_textbox(
    aes(x= 8+(5/31)-.01, y = 200,
        label = "**Aug. 5, 2022:**<br><br>Indiana passes legislation prohibiting 
    abortion from conception with few exceptions."),
    label.colour = NA, fill = panel_c,
    size = rel(2.65), color = "gray50", 
    hjust = 1, vjust = 0,  box.color = NA, width = unit(.9, "inch"))+
  
  # text and line for Sept. 15 law going into effect
  # vert line
  geom_segment(aes(x=9+(15/30), xend = 9+(15/30),
                   y = 200, yend = 1200),
               color = "gray80", linetype = "longdash", linewidth = .15)+
  # label
  geom_textbox(
    aes(x= 9+(15/30)-.01, y = 200,
        label = "**Sept. 15, 2022:**<br><br>Indiana law goes into effect."),
    label.colour = NA, fill = panel_c,
    size = rel(2.65), color = "gray50", 
    hjust = 1, vjust = 0,  box.color = NA, width = unit(.9, "inch"))+
  
  # text and line for Sept. 22 law blocked
  # vert line
  geom_segment(aes(x=9+(22/30), xend = 9+(22/30),
                   y = 200, yend = 1200),
               color = "gray80", linetype = "longdash", linewidth = .15)+
  # label
  geom_textbox(
    aes(x= 9+(22/30)+.01, y = 1110,
        label = "**Sept. 22, 2022:**<br><br>Indiana Circuit Court Judge Kelsey 
        Blake Hanlon blocks new law because it restricts bodily autonomy."),
    label.colour = NA, fill = panel_c,
    size = rel(2.65), color = "gray50", 
    hjust = 0, vjust = 1,  box.color = NA, width = unit(1.08, "inch"))+
  
  # Marion judge again blocks law
  # vert line
  geom_segment(aes(x=12+(2/31), xend = 12+(2/31),
                   y = 200, yend = 1200),
               color = "gray80", linetype = "longdash", linewidth = .15)+
  # label
  geom_textbox(
    aes(x= 12+(2/31)+.01, y = 1110,
        label = "**Dec. 2, 2022:**<br><br>Marion County Judge Heather Welch 
        blocks new law because it likely violates Indiana’s Religious Freedom 
        Restoration Act."),
    label.colour = NA, fill = panel_c,
    size = rel(2.65), color = "gray45", 
    hjust = 0, vjust = 1,  box.color = NA, width = unit(1.08, "inch"))+
  
  # monthly average indicator line
  geom_segment(aes(x=1, xend = 12,
                   y = mean(mnth$tot[mnth$year !=2022]), 
                   yend = mean(mnth$tot[mnth$year!=2022])),
               color = "gray45", linewidth = .2) +
  # monthly average label
  geom_textbox(
    aes(x= 0.5, y = mean(mnth$tot[mnth$year !=2022])),
    label = "Overall monthly average 2014-2021", 
    label.colour = NA, fill = panel_c,
    size = rel(2.1), color = "gray45", lineheight = 1.3,
    hjust = .5, box.color = NA,   width = unit(.4, "inch"))+
  # monthly average label line
  geom_segment(aes(x= .75, xend = .95, 
                   y = mean(mnth$tot[mnth$year !=2022]), 
                   yend = mean(mnth$tot[mnth$year!=2022])),
               color = "gray45", linewidth = .1)+
  
  # line for 2021 totals
  geom_line(data = mnth %>% filter(year == 2021),
            aes(month, tot, group = year), color = c_2021, size = .65)+
  # line for 2022 totals
  geom_line(data = mnth %>% filter(year == 2022),
            aes(month, tot, group = year), color = c_2022, size = .65)+
  # labels for years 
  geom_text_repel(data= mnth %>% 
                  group_by(year) %>% mutate(a_tot = sum(res)+sum(nonres)) %>% 
                  filter(month == "December"),
                  aes(12.25, tot, label = paste0(year, ":  ", prettyNum(a_tot, 
                      big.mark = ",", scientific = FALSE))),
                  #xlim = c(2021.4),
                  point.padding = .025,
                  box.padding = .2,
                  #nudge_y = -.2, 
                  nudge_x = 1.2, # .7
                  force = 1,
                  color = c_mnth$color, 
                  segment.color = "gray75",
                  size = rel(2.35), 
                  fontface = "plain", 
                  direction = "y",
                  segment.size = 0.1,
                  min.segment.length = .15,
                  hjust = .4,
                  seed = 2
  )+
  # label for July peak
  geom_textbox(
    aes(x= 5.17, y = 1135),
    label = paste0("Monthly abortions peaked in July <span style= 
    'color:",c_2022,"'>2022 </span> at 1,182."),
    label.colour = NA, fill = panel_c,
    size = rel(2.75), color = "gray30", 
    hjust = 0, box.color = NA,   width = unit(1.1, "inch"))+
  # label for year labels
  geom_textbox(
    aes(x= 12.6, y = 800),
    label = paste0("Year"),
    label.colour = NA, fill = panel_c,
    size = rel(1.95), color = "gray30", 
    hjust = 0, box.color = NA,   width = unit(1.1, "inch"))+
  # label for yearly total labels
  geom_textbox(
    aes(x= 12.95, y = 800),
    label = paste0("Total"),
    label.colour = NA, fill = panel_c,
    size = rel(1.95), color = "gray30", 
    hjust = 0, box.color = NA,   width = unit(1.1, "inch"))+
  # set y scale
  scale_y_continuous(limits = c(200, 1200),
                     breaks = c( 400, 600, 800, 1000, 1200),
                     labels = c("400", "600", "800", "1,000", "1,200"),
                     name = "Abortions")+
  # set x scale
  scale_x_discrete(
    breaks = c("January", "February", "March", "April","May", "June",
               "July", "August", "September", "October", "November","December"),
    labels = c("January", "February", "March", "April","May", "June",
               "July", "August", "September", "October", "November","December"),
    expand = expansion(add = 1.25))+
  t_theme()+
  theme(
    panel.grid.major.y = element_blank()
  )+
  labs(title = 
         paste0("Though the draft of the Dobbs decision to end federal abortion 
    protection began circulating in May 2022, the biggest impact on Indiana's <span
    style= 
    'color:",c_2022,"'>2022 </span>monthly abortion totals occurred in July when they
    jumped **",round(100*(mnth$tot[mnth$year == 2022 & mnth$month == "July"]-
    mnth$tot[mnth$year == 2021 & mnth$month == "July"])/mnth$tot[mnth$year == 2021 & 
    mnth$month=="July"]),"%** from the same month in <span style= 
    'color:",c_2021,"'>2021 </span>to a nine-year high of **1,182**. They fell to a 
    record low of **447** the following October."),
       
       caption = 
         "<br>**Data:**'www.in.gov/health/vital-records/vital-statistics/terminated-pregnancy-reports'<br>
    **Graphic:** Ted Schurter 2023")


```

### Data

```{r, data: month_yr}

datatable(mnth,
          colnames = c("Year", "Month", "Resident\nabortions", "Nonresident\nabortions", "Total\nabortions"),
          caption = htmltools::tags$caption(
            style = 'caption-side: top; text-align: center;',
            "Abortions in Indiana by year, month and residency 2014 to 2022.", style = "color:black"),
          rownames=FALSE, 
          filter = "top", 
          extensions = c('Buttons', 'FixedColumns'),
          options = list(pageLength=10, 
                         scrollX=T,
                         dom = 'Bfrtip',
                         scrollY = '250px',
                         buttons = c('csv', 'excel')
                         #columnDefs = list(list(className = 'dt-center', targets = 0:4))
          ))%>% 
  formatRound(c("tot"), digits = 0)

```

## {-}

---

#### The bulk of the boom belongs to people travelling to Indiana for abortions and who made up one in five of them in 2022. While it  was the first state to pass more restrictive abortion legislation after Dobbs, Indiana's ban didn't take effect until September 15. Kentucky, Ohio and Tennessee had trigger laws in place that outlawed abortion if federal protections were removed. An influx from those states fueled a `r formatC(100*((mnth$nonres[mnth$year == 2022 & mnth$month == "July"]-mnth$nonres[mnth$year == 2021 & mnth$month == "July"])/mnth$nonres[mnth$year == 2021 & mnth$month == "July"]), big.mark = ",")`% year-over-year increase in nonresident abortions in July and an overall annual increase of `r round(100*((sum(mnth$nonres[mnth$year == 2022])-sum(mnth$nonres[mnth$year == 2021]))/sum(mnth$nonres[mnth$year == 2021]))) `% in 2022. Resident abortions also spiked in July, but to a lesser degree, `r formatC(100*((mnth$res[mnth$year == 2022 & mnth$month == "July"]-mnth$res[mnth$year == 2021 & mnth$month == "July"])/mnth$res[mnth$year == 2021 & mnth$month == "July"]), big.mark = ",")`% higher than last year but lower than the 2022 peak of `r mnth$tot[mnth$year == 2022 & mnth$month == "March"]` recorded in March.

```{r, year by month by residency, fig.width=11, fig.height=6.5, fig.alt= "Line graph of total monthly abortions from 2014 to 2022 by residents and nonresidents showing record high numbers in July 2022 and record low numbers in October 2022 as well as a more than 900% increase in nonresident abortions from their 2014 to 2021 average level."}
# resident color  
res_c_2022 <- "#1b7837"

# nonresident color 
nres_c_2022 <- "#762a83"

# calculate monthly means
mnth_2 <- mnth %>% group_by(month) %>% arrange(year, month) %>%
  mutate(
    res_yoy = round(((res-lag(res))/lag(res))*100,2),
    n_res_yoy = round(((nonres-lag(nonres))/lag(nonres))*100,2)
  ) %>% ungroup()

# calculate year to year change and pivot to long data
mnth_lg <- mnth %>% arrange(year, month) %>% 
  mutate(
    yoy_ch = round(((tot-lag(tot))/lag(tot))*100,2) 
  ) %>% 
  pivot_longer(
    3:4, names_to = "location", values_to = "count"
  )

# calculate percent change from 2014 to 2021 averages
mnth_pc <- mnth %>% filter(year != 2022) %>% group_by(month) %>% mutate(
  mn_res = round(mean(res)),
  mn_nres = round(mean(nonres)),
  pct_d_res  = round(100*((res-mn_res)/mn_res),1),
  pct_d_nres = round(100*((nonres-mn_nres)/mn_nres),1)
) 

ggplot()+
  # major y axis grid lines
  geom_segment(aes(x = 0, xend = 12, y = 0, yend = 0), color     = "gray90", size = .25)+
  geom_segment(aes(x = 0, xend = 12, y = 200, yend = 200), color = "gray90", size = .25)+
  geom_segment(aes(x = 0, xend = 12, y = 400, yend = 400), color = "gray90", size = .25)+
  geom_segment(aes(x = 0, xend = 12, y = 600, yend = 600), color = "gray90", size = .25)+
  geom_segment(aes(x = 0, xend = 12, y = 800, yend = 800), color = "gray90", size = .25)+
  
  scale_y_continuous(limits  = c(0, max(mnth$res)),
                     breaks = c(0, 200, 400, 600, 800),
                     labels = c("0", "200", "400", "600", "800"))+
  # yearly nonres lines
  geom_line(data = mnth_lg %>% filter(location == "nonres") %>%  group_by(year) %>%  arrange(desc(count)),
            aes(month, count, group = year), color = "#8c6bb1", size = .25, alpha = .25)+
  # yearly resident lines
  geom_line(data = mnth_lg %>% filter(location == "res") %>%  group_by(year) %>%  arrange(desc(count)),
            aes(month, count, group = year), color = "#41ae76", size = .25, alpha = .25)+
  # 2022 resident line
  geom_line(data = mnth_2 %>% filter(year == 2022),
            aes(month, res, group = year),
            color = res_c_2022, linewidth = .75)+
  # 2022 nonresident line
  geom_line(data = mnth_2 %>% filter(year == 2022),
            aes(month, nonres, group = year),
            color = nres_c_2022, linewidth = .75)+
  # average nonresident line
  geom_line(data = mnth_2 %>% group_by(month) %>% filter(year != 2022) %>% mutate(mn_nr = mean(nonres)),
            aes(month, mn_nr, group = year),
            color = "gray30", linewidth = .25)+
  # average resident line
  geom_line(data = mnth_2 %>% group_by(month) %>%  filter(year != 2022) %>% mutate(mn_r = mean(res)),
            aes(month, mn_r, group = year),
            color = "gray30", linewidth = .25)+  
  
  scale_x_discrete(expand = expansion(mult = 0, add = c(0, 1.5)))+
  # label for 2022 resident abortions
  geom_textbox(aes( x = 12.15, 
                    y = mnth_2 %>% filter(year == 2022 & month == "December") %>% arrange(res) %>% pull(3)),
               label = "2022 resident abortions", label.colour = NA, fill = panel_c, size = rel(3), 
               color = res_c_2022, hjust = 0, vjust = .8,  box.color = NA, width = unit(1.25, "inch"))+
  
  # average nonresident abortions
  geom_textbox(aes( x = 12.15,
                    y = mnth_2 %>% group_by(month) %>% mutate(mn_nr = mean(nonres)) %>% filter(month == "December" & 
                                                                                                 year == 2022) %>% pull(8)),
               label = "Average nonresident abortions 2014:2021", label.colour = NA, fill = panel_c, size = rel(2.35),
               color = "grey30", hjust = 0, vjust = .55,  box.color = NA, width = unit(1.25, "inch"))+
  # average resident abortions
  geom_textbox(aes( x = 12.15,
                    y = mnth_2 %>% group_by(month) %>% mutate(mn_r = mean(res)) %>% filter(month == "December" & 
                                                                                             year == 2022) %>% pull(8)),
               label = "Average resident abortions 2014:2021", label.colour = NA, fill = panel_c, size = rel(2.35),
               color = "grey30", hjust = 0, vjust = .45,  box.color = NA, width = unit(1.25, "inch"))+
  
  # label for 2022 nonresident abortions
  geom_textbox(aes( x = 12.15,
                    y = mnth_2 %>% filter(year == 2022 & month == "December") %>% arrange(nonres) %>% pull(4)),
               label = "2022 nonresident abortions", label.colour = NA, fill = panel_c, size = rel(3),
               color = nres_c_2022, hjust = 0, vjust = .35,  box.color = NA, width = unit(1.25, "inch"))+
  # nonresident July annotation
  # label for 2022 nonresident abortions
  geom_textbox(aes( x = 6.4,
                    y = mnth_2 %>% filter(year == 2022 & month == "July") %>% arrange(nonres) %>% pull(4)),
               label = paste0("Nonresident abortions reach an all-time high of ",
                              mnth$nonres[mnth$year == 2022 & mnth$month == "July"],
                              " in July 2022, ",
                              ((mnth$nonres[mnth$year == 2022 & mnth$month == "July"]-
                                  mnth_pc$mn_nres[mnth_pc$year == 2021 & mnth_pc$month == "July"])/
                                 mnth_pc$mn_nres[mnth_pc$year == 2021 & mnth_pc$month == "July"])*100,
                              "% above average."),
               label.colour = NA, fill = panel_c, size = rel(4),
               color = nres_c_2022, hjust = 1, vjust = .85,  box.color = NA, width = unit(2.25, "inch"))+
  # label for 2022 October resident
  # label for 2022 resident October
  geom_textbox(aes( x = 10.25,
                    y = mnth_2 %>% filter(year == 2022 & month == "October") %>% arrange(res) %>% pull(3)),
               label = paste0("Resident abortions reach an all-time low in October 2022, ",
                              round(((mnth$nonres[mnth$year == 2022 & mnth$month == "October"]-
                                        mnth_pc$mn_nres[mnth_pc$year == 2021 & mnth_pc$month == "October"])/
                                       mnth_pc$mn_nres[mnth_pc$year == 2021 & mnth_pc$month == "October"])*100),
                              "% below average."),
               label.colour = NA, fill = panel_c, size = rel(2.85),
               color = res_c_2022, hjust = 0, vjust = .85,  box.color = NA, width = unit(1.25, "inch"))+
  geom_text_repel(data= mnth %>% filter (month == "January"),
                  aes(.95, res, label = year),
                  xlim = 0.3,
                  point.padding = .015,
                  box.padding = .2,
                  nudge_y = 2, 
                  #nudge_x = .2, # .7
                  force = 1,
                  color = "gray55",
                  segment.color = "gray85",
                  size = rel(2.15), 
                  fontface = "plain", 
                  direction = "y",
                  segment.size = 0.2,
                  #min.segment.length = 3.05,
                  hjust = 1,
                  seed = 2
  )+
  geom_textbox(aes(x = .03,
                   y = 110),
               label = "Each line represents a year between 2014 to 2022.", label.colour = NA, fill = panel_c, size = rel(2.15),
               color = "gray60", hjust = 0, vjust = .55,  box.color = NA, width = unit(.65, "inch"))+
  t_theme()+
  theme(
    panel.grid.major.y = element_blank(),
    
  )+
  
  labs(
    title = 
      paste0("Following the post-Dobbs enactment of abortion bans in neighboring states, 
      Indiana's
      <span style= 'color:",nres_c_2022,"'>nonresident abortions</span> surged to
      record highs in 2022, soaring ", 
             formatC(((mnth$nonres[mnth$year == 2022 & mnth$month == "July"]-
                         mnth_pc$mn_nres[mnth_pc$year == 2021 & mnth_pc$month == "July"])/
                        mnth_pc$mn_nres[mnth_pc$year == 2021 & mnth_pc$month == "July"])*100,
                     format = "f", digits = 0, big.mark = ","),
             "% past their average July level and ending ",
             round(((mnth %>% group_by(year) %>% summarise(sum(nonres)) %>% filter(year == 2022) %>% pull(2)-
                       mnth %>% filter(year != 2022) %>% group_by(year) %>% summarise(tot = sum(nonres)) %>% 
                       summarise(mean(tot)))/
                      mnth %>% filter(year != 2022) %>% group_by(year) %>% summarise(tot = sum(nonres)) %>% 
                      summarise(mean(tot)))*100),
             "% above the 2014 to 2021 annual average."),
    subtitle = 
      paste0("Monthly <span style= 'color:",res_c_2022,"'>resident abortions</span> were 
      higher than average in 2022 until September. They fell to 
       ",
             mnth$res[mnth$year == 2022 & mnth$month == "October"],
             " in October, their lowest level of the past nine years; total <span style= 'color:",res_c_2022,"'>resident abortions</span> in 2022 ended ",
             round(abs(((mnth %>% group_by(year) %>% summarise(sum(res)) %>% filter(year == 2022) %>% pull(2)-
                           mnth %>% group_by(year) %>% summarise(sum(res)) %>% filter(year == 2021) %>% pull(2))/
                          mnth %>% group_by(year) %>% summarise(sum(res)) %>% filter(year == 2021) %>% pull(2))*100)),
             "% below average."),
    caption = "<br>**Data:** 'www.in.gov/health/vital-records/vital-statistics/terminated-pregnancy-reports'<br>
  **Graphic:** Ted Schurter 2023"
  )
```

---

#### `r stringr::str_to_sentence(english::as.english(round(100*sum(nres$abortions[nres$state != "Illinois"& nres$state != "Michigan" & nres$state != "Other"])/sum(nres$abortions))))` percent of nonresidents who received an abortion in Indiana in 2022 came from Kentucky, Ohio or Tennessee, states with their own strict abortion bans. Abortions to Illinois residents in Indiana fell `r round(abs(100*((nres$abortions[nres$year==2022 & nres$state == "Illinois"]-nres$abortions[nres$year==2021 & nres$state == "Illinois"])/nres$abortions[nres$year==2021 & nres$state == "Illinois"])))`% in 2022 while those to Michigan residents remained the same. Both states allow abortions up to fetal viability. 

```{r, combo plot nonresidents, fig.width=11, fig.height=6.5, echo=FALSE, include=F}
# set colors 

# kentucky color
ky_c <-  "#fc8d59"
# ohio color
oh_c <-  "#4575b4"
# Tennessee color
tn_c <-  "#91bfdb"

s_cols <- c("Kentucky" = ky_c,
            "Illinois" = "gray",
            "Michigan" = "gray",
            "Ohio"     = oh_c,
            "Tennessee" = tn_c,
            "Other"    = "gray")
 
# import states geometry  ####

states <- states(cb=T)

# filter to neighboring states

states <- states %>% filter(NAME == "Illinois" | NAME == "Michigan" | NAME == "Kentucky" |
                              NAME == "Tennessee" | NAME == "Indiana" | NAME == "Ohio")

# order as factor

nres$state <- factor(nres$state, levels = c("Kentucky", "Ohio", "Tennessee", 
                                            "Illinois", "Michigan", "Other"))

# add colors to match bar chart #### 
states$color[states$NAME == "Kentucky"] <- ky_c 
states$color[states$NAME == "Ohio"] <- oh_c
states$color[states$NAME == "Tennessee"] <- tn_c
states$color[states$NAME == "Illinois" ] <- "gray"
states$color[states$NAME == "Michigan"] <- "gray"

#### # create combo bar chart inset map ####

# bar chart:

pt <- nres %>% group_by(state) %>% arrange(desc(abortions)) %>% ungroup() %>% 
  ggplot(aes(year, abortions, fill = state))+
  geom_col(show.legend = F)+
  scale_fill_manual(values = s_cols)+
  facet_grid(~state, scales="free_y", switch = "x")+
  scale_y_continuous(limits = c(0, 1000),
                     breaks = c(0, 250, 500, 750, 1000),
                     labels = c("0", "250", "500", "750", "1,000"))+
  t_theme()+
  theme(
    panel.grid.major.y = element_blank(),
    strip.background = element_blank(),
    strip.text = element_text(hjust = 0),
    
    plot.subtitle = element_textbox_simple(size = rel(1.1), color = "gray40", hjust = 0,
                                           lineheight = 1,margin = margin(.25, 0, .4, 0, unit = "cm"),),
  )

# map:

p1 <- ggplot()+
  geom_sf(data = states,
          fill = states$color)+
  t_theme()+
  theme(
    axis.text = element_blank(),
  )

# inset text 

state_label <- "<span style = 'color:#fc8d59;'> Kentucky</span>, <span style = 
      'color:#376bae;'>Ohio</span> and <span style = 'color:#82b6d6;'>Tennessee</span> 
      enacted near-total abortion bans after the Dobbs decision in 2022. Like Indiana, 
      <span style = 'color:#376bae;'>Ohio's </span>ban is currently blocked while 
      litigation proceeds. Illinois and Michigan retain abortion rights."

# inset text object

text1 <- ggplot(data = tibble(x = 0, y = 1, label = state_label)) +
  aes(x = x, y = y, label = label) +
  geom_textbox(
    box.color = NA,
    color = "gray25",
    fill = NA,
    width = unit(10, "cm"),
    hjust = 0,
    vjust = 1 
  ) +
  scale_x_continuous(limits = c(0, 1))+
  scale_y_continuous(limits = c(0, 1))+
  theme_void() 

# add it all up with patchwork 

combo_plot <- 
pt + inset_element(p1, 
                   left = 0.75, 
                   bottom = 0.1, 
                   right = 1, 
                   top = 1.25,
                   align_to = "plot")+
  inset_element(text1, 
                left = 0.36, 
                bottom = 0.45, 
                right = .76, 
                top = .80,
                align_to = "plot")+
  
  plot_annotation(
    theme = t_theme(),
    title =
      paste0("The number of nonresident abortions in Indiana in 2022 increased 
      dramatically following the Dobbs decision ending federal abortion protections. 
      <span style ='color:#fc8d59;'>Kentucky</span> residents received ",
             nres$abortions[nres$year == 2022 & nres$state == "Kentucky"],
             ", the most in 2022 and a ",
             round(100*((nres$abortions[nres$year == 2022 & nres$state == "Kentucky"]-
                           nres$abortions[nres$year == 2021 & nres$state == "Kentucky"])/
                          nres$abortions[nres$year == 2021 & nres$state == "Kentucky"])),
             "% increase from 2021. <span style ='color:#4575b4;'>Ohio's</span> totals increased ",
             formatC(round(100*((nres$abortions[nres$year == 2022 & nres$state == "Ohio"]-
                                   nres$abortions[nres$year == 2021 & nres$state == "Ohio"])/
                                  nres$abortions[nres$year == 2021 & nres$state == "Ohio"])),big.mark = ","),
             "% from 2021.")
  )
```

## {.tabset}

### Chart

```{r, display combo plot, fig.width=11, fig.height=6, echo=FALSE, fig.alt= "Bar chart of nonresident abortion totals for states near Indiana that had residents travel here for an abortion showing Kentucky with a 260% increase from 2021 and Ohio with a 1,428% increase. Chart also include inset map of states color coded according to the existence of post-Dobbs abortion bans."}
combo_plot
```


### Data

```{r, data: nonres by year, state}

datatable(nres,
          colnames = c("Year", "State", "Abortions"),
          caption = htmltools::tags$caption(
            style = 'caption-side: top; text-align: center;',
            "Nonresident abortions in Indiana by year and state 2019 to 2022.", style = "color:black"),
          rownames=FALSE, 
          filter = "top", 
          extensions = c('Buttons', 'FixedColumns'),
          options = list(pageLength=10, 
                         scrollX=T,
                         dom = 'Bfrtip',
                         scrollY = '250px',
                         buttons = c('csv', 'excel')
                         #columnDefs = list(list(className = 'dt-center', targets = 0:4))
          ))

```

## {-}

---
<br><br>

###### Data and analysis workflow available at [https://github.com/tedschurter/tpr_2022](https://github.com/tedschurter/tpr_2022/)

###### A previous analysis of Indiana abortion data from 2014 to 2021 can be found [here.](https://tedschurter.github.io/indiana_abortion/)



